Reinforcement-Learning Windturbine Controller

### 2019-12-09 paper/paper.tex
Abgabe

\documentclass[hyperref,beleg]{cgvpub}

@@

\betreuer{Dr. Dmitrij Schlesinger}
\bibfiles{literatur}
\problem{
  \section{Background:}
  Here at the HFI Experimental fluid mechanics group, we have developed an
  open source project called QBlade. QBlade is a simulation tool used for
  testing wind turbines in the hostile environment that they normally operate.
  We normally tackle problems of aerodynamic or structural optimization but
  we have also a research focus on the development of the control systems of
  the wind turbines. We currently have a research effort looking at developing
  cluster-based controllers building on the work of Professor Bernd Noack who
  is a guest professor at our group. In the last year or so (Nair, A. G., Yeh, C.-
  A., Kaiser, E., Noack, B. R., Brunton, S. L., \& Taira, K. (2018). Cluster-based
  feedback control of turbulent post-stall separated flows. Journal of Physics
  Fluid Dynamics, (M), 1-32. Retrieved from http://arxiv.org/abs/1809.07220).
  AI projects such as openAI have enabled he rapid development of neural
  network in the field of control using reinforcement learning. The goal of this
  project is to use QBlade as a wind turbine simulator and attempt to control
  the pitch and rotor speed in a way that doesn’t cause the wind turbine to
  shatter but instead to yield energy, i.e. reward and death condition. This first
  stage of work should be considered as exploratory but will hopefully open up
  avenues of controlling active flow control elements such as flaps.

  \section{Tasks}
  The major tasks of the project are as follows:
  \begin{itemize}
    \item Build up and interface between QBlade and python the model code so that an external code can run as a controller within a QBlade simulation.
    \item Gain a rough understanding of the mechanics of wind turbines and their controllers.
    \item Research reinforcement learning methods suitable for use as a windturbine controller and perform a literature review on these approaches.
    \item Create a reinforcement learning agent which uses the Qblade interface for controllers to control a windturbine. 
      \begin{itemize}
        \item Inputs to the agent could be defined by the standartized controller input format to Nordex turbines, which consists of 39 real-valued sensor-inputs. However, initial tests can be conducted with whichever inputs are easiest to tackle. If required, further hidden state from the simulation can be exported to enrich data quality. If aiming for industrial quality, more inputs and also sensor faults could be optionally incorporated.
        \item Outputs are in a minimum version pitch angles for the 3 blades and turbine torque. Optionally the agent should be able to control active element such as flaps on the blades.
      \end{itemize}
    \item Optimize the agent to deliver maximum energy yield.
    \item Optimize under respect of certain boundary conditions (maximum pitch acceleration, maximum power, maximum blade load, blade touching the tower) and optionally other boundary conditions like long term turbine wear.
    \item If necessary for the training process, scale the simulation to run at a larger scale.
    \item Implement and attempt to get the agent to perform something close to sensible control of the wind turbine. Optionally evaluate the results against existing controllers and try to outperform them.
    \item Optionally, create a conference paper, poster or blog post etc.. on the results. 
  \end{itemize}
}

\acknowledgments{At first thanks to my advisors Mattew Lennie and Dmitrij Shlezinger for their support. Also, I guess, thanks David Silver - 7 of my most important papers were from you. Thanks to my laptop for almost dying, but then surviving. And big thanks to the mighty god of windturbines for letting me crush several thousand windturbines for science.}

@@

Recent advancements in reinforcment learning have managed to tackle more and more complex problems, like StarCraft or Go. The range of topics, RL is applicable to, increases. However, so far, these advancements were only partially applied to the area of windturbine pitch and torque control. Current state of the art linear controllers are performing well in maintaining turbine control, and the optimization margin for these is small, as they achieve the theoretical maximum energy output up to rated windspeed and manage to keep the turbine intact above rated windspeeds. However, these simple controllers can only deal with a limited amount of data inputs and are preventing turbine damages. We are evaluating the use of an actor-critic reinforcement learning algorithm to solve continous control of a windturbine, hoping to lay a foundational work for later improvements. We find that utilizing a wildly proven continous control algorithm does not solve our windturbine problem. Finally, we demonstrate that this control is possible, however we do not achieve results outperforming industry controllers.

@@

With the issue of global warming, we are facing maybe the biggest challenge to our generation. Though wheather models have predicted the effect since the 1980s, only in the recent years this topic has been brought into societal focus. CO2 is seen as the main cause for a tendency in the global weather to heat up. Since our industrial age, humankind has been releasing CO2 into the athmosphere in large scales. If however we continue doing so the way we are doing now, we will likely increase global temperature by a degree which will make living on this planet difficult. With the Paris agreement, the United Nations decided to limit global warming to 1.5 degrees, for which massive reductions in CO2 emissions are necessary. As energy production holds one of the biggest parts in the global CO2 emissions, we need to replace CO2 emitting power plants. For this, renewable energies have played a major role, and as a big part of this also windpower.

Winds are a direct result of sunshine. The sun heats different parts of the earth to different temperatures, as the surface of the earth isn't uniform. The air over hotter regions such as close to the equator tends to rise, while on mountains and the poles it tends to fall. Combine this with a vast and complex interplay of humidities, Coriolis effects, mountains, cloud formation and enough effects to fill an entire branch of science and you will get wind. Windturbines operate by drawing energy from this weather phenomenon. Large rotors placed in windy regions which are connected to a generator yield electrical power. As weather is a complex phenomenon, wind tends to come from different directions with different speeds. This makes drawing energy from it more complex, or less efficient. If we were to draw energy from both storms and light breezes with the same turbine, we would not be efficient. Thus, modern windturbines are built to be adjustable. With these adjustments, a turbine can operate sensibly over a broader range of wind conditions. However, somehow, these adjustments need to be automatically adapted to the wind condition. This is the task of the turbine controller. We are attempting to learn such a controller.

Machine learning has developped immensly over the last years. Benefitting from Moore's law, huge amounts of computational power can be used to train more and more complex models. Machine learning is generally split into three big branches. Supervised learning, unsupervised learning and reinforcement learning. In supervised learning, a function approximator is fitted to model properties of a dataset. In unsupervised learning, properties in a dataset are measured, compared or grouped. In reinforcement learning, the dataset is explored interactively. We are trying to learn a controller for a windturbine, so in theory we could use both supervised an reinforcement learning. If we were to use supervised learning, we would need some precomputed data, which could only be generated by already running a windturbine with a controller and measuring the data. As we however want to learn this controller, it is not present at the start of supervised learning, and imitating known controllers could at best get exactly as good as the known controller. So we are using reinforcement learning.

@@

\section{Windturbine control}
% Quote wind-energy handbook, talk about how currently wind turbines are controlled.
There are two mayor types of wind turbine designs, \ac{HAWT} and \ac{VAWT}, which differ by their rotation axis. In this work, we will only look at \ac{HAWT} and all mentions of windturbines mean \ac{HAWT} type turbines. Such a turbine is made up by 3 big components, a tower on which a nacelle is rested which itself has a rotor in front. 
The joint between tower and nacelle allows for rotation to turn the rotor into the wind. Except for recent ideas \cite{howlandWindFarmPower2019}, the best value for the rotational direction was to always yaw the rotor directly into the wind, and for the sake of simplicity we will omit this control parameter from our simulation and leave it at 0 degrees with wind facing straight onto the rotor.
In the nacelle there is, most importantly, an electrical generator, which can be given a specific torque. It will then slow down the rotor with that torque and at the same time generate energy proportional to rotor speed and torque. This torque is one of the two more important control parameters, together with blade pitch. The blades are designed to operate on maximal aerodynamic efficiency when they are pitched at 0 degrees, increasing the pitch angle will (except for possible slight improvements in the first few degrees) reduce aerodynamic efficiency of the rotor.

\begin{figure}
  \centering
  \includegraphics{images/NREL-5MW.png}
  \caption{Control curve of the NREL 5MW turbine, from \cite{jonkmanDefinition5MWReference2009}}
  \label{fig:nrel5mw}
\end{figure}

@@

Usually this control is implemented by two PID controllers, which are hand-designed as described in \cite[sec 8.4]{burtonWindEnergyHandbook2011}. PID-Controllers in general have one real-valued input over time and deliver an output. This is done by calculating an error term between the input and a desired input. The output is calculated as a mixture of the error itself (P - proportional to the error), integrating the error over the last few timesteps (I - integral of the error) and calculating a derivative to the last timestep (D - derivative). The output is the sum of these 3 terms each factorized with a constant factor.
However back to windturbines. Usually, we implement two PID controllers to control a turbine, a torque and a pitch controller. Below rated windspeed, the controller for torque is active, above rated the one for pitch. The parameters to those controllers can be calculated according to the laws of control theory. Sometimes, this strong split into two operational ranges is relaxed a bit, as in the 5MW controller, the pitch part counteracts resonances also below rated speed. From this simple concept, the resulting controllers work reasonably close to theoretical maximum already. Our reference turbine has a peak power coefficient of 0.482 - meaning that at optimal wind speed, 48\% of the energy of the wind is converted into electricity. In fact, the theoretical maximum is given by the Betz limit of 59.3\% and if we account for electrical and frictional losses, the realistic maximum is even lower. So there is not much potential to be alleviated. Also, only a fraction of this potential lies in the responsibility of the control circuit. Blade design has so far been the major area for improvements.

@@

Also, the outputs get more complex. Back in the early days of wind turbines, you could control 3 variables - pitch, torque and yaw. Nowadays, we can control each of the blades individually and with a quick enough response time to wiggle the blades around within one single rotor rotation. Research on adding active control elements to the blades such as flaps is happening, which would add many more control parameters. You might have seen flaps on an aircraft wing, extending the length of the wing or adding gaps at takeoff and landing to account for the low wind speeds hitting the aircraft wing in these flight scenarios. This could also benefit wind turbines at low wind speeds, however building a good controller for them is challenging. Tackling all this with \ac{SISO} is possible, but a lot of manual work.

Next, the inputs also get more complex. It is getting cheaper to install high-precision sensors, and many of them. Wind measurement techniques such as LIDAR based aenometers are able to efficiently and precisely predict incoming wind situations. Load measurements can be done in real-time across a turbine blade and in theory, it would be possible to incorporate these into control systems. However, accomodating all these input parameters in \ac{SISO} systems is difficult as we would have to hand-model each of the inputs and how exactly it ends up in the output.

We hope to lay a foundational step in solving these issues by using an end-to-end machine learning algorithm which is capable of learning a wind turbine control scenario with many inputs and outputs. Also, we must admit that we are generally interested in playing around with reinforcement learning, it's fucking cool.

\section{QBlade}
% Quickly introduce qblade as a simulation tool
Our data source in this work is the open-source simulation tool QBlade \cite{martenQBLADEOPENSOURCE2013} developed at TU Berlin. QBlade is an accessible and performant tool with the primary purpose of designing and simulating wind turbines in a graphical user interface. Its simulation results are on par with current state-of-the-art proprietary simulation tools and it yields good computational efficiency. It uses an algorithm which approximates the 3D blade structures with 2D structures and corrects the results through several error terms. Though a full CFD simulation would yield higher accuracy, our computational resources don't allow for that. To be used in reinforcement learning, we designed an interface, over which the QBlade simulation can be embedded into an external environment. Most machine-learning frameworks are written in python, so we decided to compile QBlade into library format and expose the most fundamental functions to allow it to communicate with any programming language that can load libraries. As python has a ctypes interface to load c-code, we could link a python agent to the QBlade C++ environment.

QBlade allows different simulation scenarios, for our testing we decided to only use the NREL 5MW \cite{jonkmanDefinition5MWReference2009} turbine with the default structural model and using all implemented correction mechanisms to achieve the most realistic data possible. As reference data to this turbine is easily available and publicly published by NREL, this allows us for good cross-validation.

The interface is made of 7 functions, which allow for loading a project, resetting the simulation, setting controller inputs and getting environment measurements and advancing the simulation a timestep. The observations returned in getControlVars match those that are visible to standart industry controllers, which in turn are modelled after what can be measured on a real windturbine. We have 23 observations and 5 actions.

After a while of experimenting with the unrestricted simulation, we added some bounds to our control parameters and hid some observations and actions. Restrictions to our final version are described in section \ref{sec:alg_qblade}.

@@

At first, we need to do some assumptions on our environment $\varepsilon$. The concept environment encapsulates the reality or simulation that the reinforcement learning algorithm works in. This could be an actual cyber-physical system or a simulation. The first assumption to this environment is that this system operates over time and we can discretize time into timesteps. Our environment here is the simulated wind turbine, but it could be a computer game \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, a physics simulation\cite{brockmanOpenAIGym2016} or even an actual existing wind turbine \cite{kolterDesignAnalysisLearning2012}. 
In every timestep, the environment supplies us with an observation \(x\) and a scalar reward \(r\) for taking an action \(a\) in that state. In theory, the full trajectory \((x_0, a_0, ..., x_t)\) could be needed to describe the full state of the simulation at timestep t \(s_t\). However, the algorithms we looked at assume that the state progressions can be modelled as a \ac{MDP} of observations, as in that what will happen at timestep t+1 is only dependent on the observation at timestep t, not also on for example t-10. In \ac{MDP}, the state transition probability \(p(x_{t+1}|x_t, a_t)\) is fully descriptive, as in it is equal to \(p(x_{t+1}|x_0, a_0, ..., x_t, a_t\). Because we assume the observation \(x\) to be fully descriptive, we will use the term state as equivalent to observation \(s_t = x_t\). In reality, the observed state doesn't have to represent the entirety of the state of the simulation, in fact it usually is only a small subset. For our wind turbine example, the observed state is limited to what can be measured with sensors on a wind turbine, in an arcade game it could be the pixel output or in a physical simulation it could be some key values in the simulation. We denote the probability of a state transition to a specific $s'$ as $p(s'|s,a;\varepsilon)$ and the distribution over all states $s'$ as $P_{s' \sim \varepsilon | s,a}$ and when $s,a$ is clear, we omit it.

@@

In every timestep, the agent supplies an action \(a\). This action could be button presses in an arcade game or blade pitch in a windturbine simulation. The mission of the agent is to pick actions which maximize cumulative reward. In other words, it should steer the environment to achieve best performance. Let's write down this mission. We want to maximize:

\begin{equation}
J = \mathbb{E}r
\end{equation}

@@

Additionally to the terms defined by the environment, we introduce some terms which are common to reinforcement learning
\begin{itemize}
  \item \textbf{Policy} A function that maps from states to actions. There are deterministic policies $\mu: S -> A$ and stochastic policies $\pi: S -> P(A)$. If we approximate a policy by a function approximator parametrized with $\theta$, we write $\pi(s;\theta)$ or omit the parameters when clear.
  \item \textbf{Value} A value with respect to a policy and a starting state s gives the accumulative reward of following that policy from state s until infinitely in the future. $V^{\pi}(s_t) = \mathbb{E}_{s_i \sim \varepsilon, a_i \sim \pi, i>t} r(s_i, a_i) $. As we have a stochastic environment and policy, we have to add an expectation term over actions taken and states resulting from those actions. It is the expected accumulative reward of following that policy. Furthermore a value can be discounted or undiscounted. Discounting means we multiply a discounting factor $\gamma \in [0,1]$ to rewards in the fashion of $r_0 \gamma^0 + r_1 \gamma^1 + r_2 \gamma^2 ...$, which results in distant future rewards being weighted less than closer ones.
  \item \textbf{Return} If we measured a value function, that would be the return. Obviously we can't measure infinitely, so a return is only defined on a finite trajectory of state-transitions, but otherwise is equal to a value.
  \item \textbf{Optimal policy} The theoretical construct of a policy that always takes the best action with respect to accumulative reward. Though we rarely know this policy, there must always be at least one policy which gives the highest value of all possible policies in a finite \ac{MDP}
  \item \textbf{Q-Value} The reward of taking action a in state s plus the value following a policy after that $Q^{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} V^{\pi}(s')$. When following the optimal policy, we simply write $Q^*(s,a)$. Usually, our Q functions are discounted. Note that we can also define Values on Q-Values: $V^{\pi}(s) = \mathbb{E}_{a \sim \pi} Q(s, a)$. If we approximate Q by a function approximator parametrized with $\theta$, we write $Q(s, a;\theta)$ or omit it in case it is clear.
  \item \textbf{Actor-Critic} Actor is just another name for a learnt policy and critic just another name for an estimated Q-Value function. These two frequently appear together, if that happens we are using an actor-critic algorithm.
  \item \textbf{State-density} According to \cite{suttonPolicyGradientMethods}, $d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$ describes how likely it is to be in state s when following a policy $\pi$ forever. If we for example have a policy which plays go as well as the authors of this paper, states with a lot of enemy stones and very few own stones might be very likely, and a state without enemy stones at all would be very unlikely. Also according to \cite{suttonPolicyGradientMethods} this distribution is independent of the starting state $s_0$ and only dependent on the policy and of course the environment. In later papers, commonly a discounted version of this is used. $\rho^{\pi}(s)$ weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$. We will omit a precise definition here, you can look it up at \cite[Section 2.1]{silverDeterministicPolicyGradient2013}.
\end{itemize}

>>>>

\documentclass[hyperref,final,beleg]{cgvpub}

@@

\betreuer{Dr. Dmitrij Schlesinger, PhD. Matthew Lennie}
\bibfiles{literatur}
\problem{
  The following problem formulation is an exact copy from what M. Lennie drafted at the beginning of the work and the only part of this extract which is not formuluated by N. Westerbeck.

  \textbf{Problem}

  Here at the HFI Experimental fluid mechanics group, we have developed an
  open source project called QBlade. QBlade is a simulation tool used for
  testing wind turbines in the hostile environment that they normally operate.
  We normally tackle problems of aerodynamic or structural optimization but
  we have also a research focus on the development of the control systems of
  the wind turbines. We currently have a research effort looking at developing
  cluster-based controllers building on the work of Professor Bernd Noack who
  is a guest professor at our group. In the last year or so (Nair, A. G., Yeh, C.-
  A., Kaiser, E., Noack, B. R., Brunton, S. L., \& Taira, K. (2018). Cluster-based
  feedback control of turbulent post-stall separated flows. Journal of Physics
  Fluid Dynamics, (M), 1-32. Retrieved from http://arxiv.org/abs/1809.07220).
  AI projects such as openAI have enabled he rapid development of neural
  network in the field of control using reinforcement learning. The goal of this
  project is to use QBlade as a wind turbine simulator and attempt to control
  the pitch and rotor speed in a way that doesn’t cause the wind turbine to
  shatter but instead to yield energy, i.e. reward and death condition. This first
  stage of work should be considered as exploratory but will hopefully open up
  avenues of controlling active flow control elements such as flaps.

  \textbf{Tasks}

  The major tasks of the project are as follows:
  \begin{itemize}
    \item Build up and interface between QBlade and python the model code so that an external code can run as a controller within a QBlade simulation.
    \item Gain a rough understanding of the mechanics of wind turbines and their controllers.
    \item Research reinforcement learning methods suitable for use as a windturbine controller and perform a literature review on these approaches.
    \item Create a reinforcement learning agent which uses the Qblade interface for controllers to control a windturbine. 
      \begin{itemize}
        \item Inputs to the agent could be defined by the standartized controller input format to Nordex turbines, which consists of 39 real-valued sensor-inputs. However, initial tests can be conducted with whichever inputs are easiest to tackle. If required, further hidden state from the simulation can be exported to enrich data quality. If aiming for industrial quality, more inputs and also sensor faults could be optionally incorporated.
        \item Outputs are in a minimum version pitch angles for the 3 blades and turbine torque. Optionally the agent should be able to control active element such as flaps on the blades.
      \end{itemize}
    \item Optimize the agent to deliver maximum energy yield.
    \item Optimize under respect of certain boundary conditions (maximum pitch acceleration, maximum power, maximum blade load, blade touching the tower) and optionally other boundary conditions like long term turbine wear.
    \item If necessary for the training process, scale the simulation to run at a larger scale.
    \item Implement and attempt to get the agent to perform something close to sensible control of the wind turbine. Optionally evaluate the results against existing controllers and try to outperform them.
    \item Optionally, create a conference paper, poster or blog post etc.. on the results. 
  \end{itemize}
}

\acknowledgments{At first thanks to my advisors Matthew Lennie and Dmitrij Shlezinger for all the support they gave in the process of writing this thesis. Thanks to David Marten for helping with QBlade. Also, I guess, thanks David Silver - 7 of my most important papers were from you. Thanks to my laptop for almost dying, but then surviving. And big thanks to the mighty god of windturbines for letting me crush several thousand windturbines for science.}

@@

Recent advancements in \ac{RL} have managed to tackle more and more complex problems, like StarCraft or Go. The range of topics, \ac{RL} is applicable to, increases. However, so far, these advancements were only partially applied to the area of windturbine pitch and torque control. Current state of the art linear controllers are performing well in maintaining turbine control, and the optimization margin for these is small, as they achieve near theoretical maximum energy output. However, these simple controllers can only deal with a limited amount of data inputs and are struggling to prevent long-term turbine damages. We are evaluating the use of DDPG as an actor-critic reinforcement learning algorithm to solve continuous control of a windturbine, hoping to lay a foundational work for later improvements. We find that utilizing the unmodified algorithm does not solve our windturbine problem, thus propose to extend DDPG with a range of known and novel methods, including prioritized experience replay, death handling and normalization. Finally, we demonstrate that windturbine control is possible, however we do not achieve results outperforming industry controllers.

@@

With the issue of global warming, we are facing maybe the biggest challenge to our generation. Though weather models have predicted the effect since the 1980s, only in the recent years this topic has been brought into societal focus. CO2 is seen as one of the main cause for a tendency in the global weather to heat up. Since our industrial age, humankind has been releasing CO2 into the atmosphere in large scales. If however we continue doing so the way we are doing now, we will likely increase global temperatures by roughly 3 degrees which will make living on this planet difficult. With the Paris agreement, the United Nations decided to limit global warming to 1.5 degrees, for which massive reductions in CO2 emissions are necessary. As energy production holds one of the biggest parts in the global CO2 emissions, we need to replace CO2 emitting power plants. For this, renewable energies have played a major role, and as a big part of this also windpower.

Winds are a direct result of sunshine. The sun heats different parts of the earth to different temperatures, as the surface of the earth isn't uniform. The air over hotter regions such as close to the equator tends to rise, while over oceans and the poles it tends to fall. Combine this with a vast and complex interplay of humidities, Coriolis effects, mountains, cloud formation and enough co-dependence to fill an entire branch of science and you will get wind. Windturbines operate by drawing energy from this weather phenomenon. Large rotors placed in windy regions which are connected to a generator yield electrical power. As weather is a complex phenomenon, wind tends to come from different directions with different speeds. This makes drawing energy from it more complex, or less efficient. If we were to draw energy from both storms and light breezes with the same turbine, we would not be efficient as the turbine can only draw so much energy that it survives the storm, which projected on the light breeze is very little energy yield. Thus, modern windturbines are built to be adjustable. With these adjustments, a turbine can operate sensibly over a broader range of wind conditions. However, somehow, these adjustments need to be automatically adapted to the inflowing wind. This is the task of the turbine controller. We are attempting to learn such a controller.

Machine learning has developed immensely over the last years. Benefiting from Moore's law, huge amounts of computational power can be used to train more and more complex models. Machine learning is generally split into three big branches. Supervised learning, unsupervised learning and reinforcement learning. In supervised learning, a function approximator is fitted to model properties of a dataset. In unsupervised learning, properties in a dataset are measured, compared or grouped. In reinforcement learning, the dataset is an environment and it is explored interactively. We are trying to learn a controller for a windturbine, so in theory we could use both supervised an reinforcement learning. If we were to use supervised learning, we would need some precomputed data, which could only be generated by already running a windturbine with a controller and measuring the data. As we however want to learn this controller, it is not present at the start of supervised learning, and imitating known controllers could at best get exactly as good as the known controller. So we are using reinforcement learning.

@@

\section{Windturbine control}
% Quote wind-energy handbook, talk about how currently wind turbines are controlled.
There are two major types of wind turbine designs, \ac{HAWT} and \ac{VAWT}, which differ by their rotation axis. In this work, we will only look at \ac{HAWT} as these are more commonly used, in our work all mentions of windturbines mean \ac{HAWT} type turbines. Such a turbine is made up by 3 big components, a tower on which a nacelle is rested which itself has a rotor in front. 
The joint between tower and nacelle allows for rotation to turn the rotor into the wind, and the angle between inflowing wind and nacelle orientation is called yaw angle. Except for recent ideas \cite{howlandWindFarmPower2019}, the best value for the rotational direction was to always yaw the rotor directly into the wind, and for the sake of simplicity we will omit this control parameter from our simulation and leave it at 0 degrees with wind facing straight onto the rotor.
In the nacelle there is, most importantly, an electrical generator, which can be given a specific torque. Torque is a word for the amount of rotational force, here in opposite direction to rotation. This force will slow down the rotor and at the same time generate energy. If generator torque is approximately equal to the torque created by wind, the two cancel out and leave the rotor at constant rotational speed, which is what we want to achieve. This torque is one of the two more important control parameters, together with blade pitch. The blades are designed to operate on maximal aerodynamic efficiency when they are pitched close to 0 degrees, increasing the pitch angle will (except for possible slight improvements in the first few degrees) reduce aerodynamic efficiency of the rotor. The main use case of blade pitch is to protect the turbine from damage, as high wind speeds can generate higher loads than the turbine could handle. Our generator has a maximum torque limit, and if the inflowing wind generates more than that torque, the turbine would spool up faster than what it can handle. We want to pitch the blades as little as possible so we can maintain maximum generator torque.
\begin{figure}
  \centering
  \includegraphics{images/NREL-5MW.png}
  \caption{Control curve of the NREL 5MW turbine, from \cite{jonkmanDefinition5MWReference2009}}
  \label{fig:nrel5mw}
\end{figure}

@@

Usually this control is implemented by two PID controllers, which are hand-designed as described in \cite[sec 8.4]{burtonWindEnergyHandbook2011}. PID-Controllers in general have one real-valued input over time and deliver an output. This is done by calculating an error term between the measured input and a desired input. The output is calculated as a mixture of the error itself (P - proportional to the error), integrating the error over the last few time-steps (I - integral of the error) and calculating a derivative to the last time-step (D - derivative). The output is the sum of these 3 terms each factorized with a constant factor. There are well proven formulas how to choose these parameters and also some adaptive methods, but control theory usually chooses these parameters by hand.
However back to windturbines. Usually, we implement two PID controllers to control a turbine, a torque and a pitch controller. Below rated windspeed, the controller for torque is active, above rated the one for pitch. The parameters to those controllers can be calculated according to the laws of control theory. Sometimes, this strong split into two operational ranges is relaxed a bit, as in the 5MW controller, the pitch part counteracts resonances also below rated speed. From this simple concept, the resulting controllers work reasonably close to theoretical maximum already. Our reference turbine has a peak power coefficient of 0.482 - meaning that at optimal wind speed, 48\% of the energy of the wind is converted into electricity. In fact, the theoretical maximum is given by the Betz limit of 59.3\% and if we account for unavoidable electrical and frictional losses, the realistic maximum is even lower. So there is not much potential to be alleviated. Also, only a fraction of this potential lies in the responsibility of the control circuit. Blade design has so far been the major area for improvements.

@@

Also, the outputs get more complex. Back in the early days of wind turbines, you could control 3 variables - pitch, torque and yaw. Nowadays, we can control each of the blades individually and with a quick enough response time to wiggle the blades around within one single rotor rotation. Research on adding active control elements to the blades such as flaps is happening, which would add many more control parameters. You might have seen flaps on an aircraft wing, extending the length of the wing or adding gaps at takeoff and landing to account for the low wind speeds hitting the aircraft wing in these flight scenarios. This could also benefit wind turbines either for higher efficiency on low wind speeds or to furthermore reduce blade loads. However building a good controller for them is challenging due to the high number of interplaying parameters. Tackling all this with \ac{SISO} is possible, but a lot of manual work.

Next, the inputs also get more complex. It is getting cheaper to install high-precision sensors, and many of them. Wind measurement techniques such as LIDAR based anemometers are able to efficiently and precisely predict incoming wind situations. Load measurements can be done in real-time across a turbine blade and in theory, it would be possible to incorporate these into control systems. However, accommodating all these input parameters in \ac{SISO} systems is difficult as we would have to hand-model each of the inputs and how exactly it ends up in the output.

We hope to lay a foundational step in solving these issues by using an end-to-end reinforcement learning algorithm which is capable of learning a wind turbine control scenario with many inputs and outputs. Also, we must admit that we are generally interested in playing around with reinforcement learning, it's fucking cool.

\section{QBlade}
% Quickly introduce QBlade as a simulation tool
Our data source in this work is the open-source simulation tool QBlade \cite{martenQBLADEOPENSOURCE2013} developed at TU Berlin. QBlade is an accessible and performant tool with the primary purpose of designing and simulating wind turbines in a graphical user interface. Its simulation results are on par with current state-of-the-art proprietary simulation tools and it yields good computational efficiency. It uses an algorithm which approximates the 3D blade structures with 2D structures and corrects the results through several error terms. Though a full 3D CFD simulation would yield higher accuracy, the increase in computational power needed doesn't justify the accuracy improvement yet. To be used in reinforcement learning, we designed an interface, over which the QBlade simulation can be embedded into an external environment. Most machine-learning frameworks are written in python, so we decided to compile QBlade into library format and expose the most fundamental functions to allow it to communicate with any programming language that can load libraries. During this work, David Marten of TU Berlin was of great help, implementing all necessary changes to QBlade so it can be compiled as a library, thank you for this! As python has a ctypes interface to load c-code, we could link a python agent to the QBlade C++ environment.

QBlade allows different simulation scenarios, for our testing we decided to only use the NREL 5MW \cite{jonkmanDefinition5MWReference2009} turbine with the default structural model and using all implemented correction mechanisms to achieve the most realistic data possible. As reference data to this turbine is easily available and publicly published by NREL, this allows for good cross-validation.

The interface is made of 7 functions, which allow for loading a project, resetting the simulation, setting controller inputs and getting environment measurements and advancing the simulation a time-step. The observations returned in getControlVars match those that are visible to standart industry controllers, which in turn are modeled after what can be measured on a real windturbine. We have 23 observations and 5 actions.

After a while of experimenting with the unrestricted simulation, we added some bounds to our control parameters and hid some observations and actions. Restrictions to our final version are described in section \ref{sec:alg_QBlade}.

@@

At first, we need to do some assumptions on our environment $\varepsilon$. The concept environment encapsulates the reality or simulation that the reinforcement learning algorithm works in. This could be an actual cyber-physical system or a simulation. The first assumption to this environment is that this system operates over time and we can discretize time into time-steps. Our environment here is the simulated wind turbine, but it could be a computer game \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, a physics simulation\cite{brockmanOpenAIGym2016} or even an actual existing wind turbine \cite{kolterDesignAnalysisLearning2012}. 
In every time-step, the environment supplies us with an observation \(x\) and a scalar reward \(r\) for taking an action \(a\) in that state. In theory, the full trajectory \((x_0, a_0, ..., x_t)\) could be needed to describe the full state of the simulation at time-step t \(s_t\). However, the algorithms we looked at assume that the state progressions can be modeled as a \ac{MDP} of observations, as in what will happen at time-step t+1 is only dependent on the observation at time-step t, not also on for example t-10. In other words, in \ac{MDP}, the state transition probability \(p(x_{t+1}|x_t, a_t)\) is fully descriptive, as in it is equal to \(p(x_{t+1}|x_0, a_0, ..., x_t, a_t)\). Because we assume the observation \(x\) to be fully descriptive, we will use the term state as equivalent to observation \(s_t = x_t\). In reality, the observed state doesn't have to represent the entirety of the state of the simulation, in fact it usually is only a small subset. For our wind turbine example, the observed state is limited to what can be measured with sensors on a wind turbine, in an arcade game it could be the pixel output or in a physical simulation it could be some key values in the simulation. We denote the probability of a state transition to a specific $s'$ as $p(s'|s,a;\varepsilon)$ and the distribution over all states $s'$ as $P_{s' \sim \varepsilon | s,a}$ and when $s,a$ is clear, we omit it.

@@

In every time-step, the agent supplies an action \(a\). This action could be button presses in an arcade game or blade pitch in a windturbine simulation. The mission of the agent is to pick actions which maximize cumulative reward. In other words, it should steer the environment to achieve best performance. Let's write down this mission. We want to maximize:

\begin{equation}
J = \mathbb{E}[r]
\end{equation}

@@

Additionally to the terms defined by the environment, we introduce some terms which are common to reinforcement learning

\begin{itemize}
  \item \textbf{Policy} A function that maps from states to actions. There are deterministic policies $\mu: S \rightarrow A$ and stochastic policies $\pi: S \rightarrow P(A)$. If we approximate a policy by a function approximator parametrized with $\theta$, we write $\pi(s;\theta)$ or omit the parameters when clear. We sometimes call a learnt policy \textit{actor}
  \item \textbf{State-density} We define a function similar to \cite{suttonPolicyGradientMethods} which describes how likely it is to be in state $s$ when following a policy $\pi$ forever. $d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$. If we for example have a policy which plays Go as well as the author of this paper, states with a lot of enemy stones and very few own stones might happen more often than others, and a state without any enemy stones would be very unlikely be seen at all. Also according to \cite{suttonPolicyGradientMethods} this distribution is independent of the starting state $s_0$ and only dependent on the policy and of course the environment. In later papers, commonly a discounted version of this is used. $\rho^{\pi}(s)$ weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$. We will omit a precise definition here, you can look it up at \cite[Section 2.1]{silverDeterministicPolicyGradient2013}.
  \item \textbf{Value} A value with respect to a policy and a starting state s gives the accumulative reward of following that policy from state s until infinitely in the future. $V^{\pi}(s_t) = \mathbb{E}_{s_i \sim \rho^{\pi}, a_i \sim \pi, i>t} r(s_i, a_i) = \int_{s_i, i>t} \rho^{\pi}(s_i) \int_{a_i, i>t} \pi(a_i) r(s_i, a_i) $. As we have a stochastic environment and policy, we have to add an expectation term over actions taken and states resulting from those actions. It is the expected accumulative reward of following that policy. Furthermore a value can be discounted or undiscounted. Discounting means we multiply a discounting factor $\gamma \in [0,1]$ to rewards in the fashion of $r_0 \gamma^0 + r_1 \gamma^1 + r_2 \gamma^2 ...$, which results in distant future rewards being weighted less than closer ones.
  \item \textbf{Return} If we measured a value function, that would be the return. Obviously we can't measure infinitely, so a return is only defined on a finite trajectory of state-transitions that we can actually observe, but otherwise is equal to a value.
  \item \textbf{Optimal policy} The theoretical construct of a policy that always takes the best action with respect to accumulative reward is called optimal policy. Though we rarely know this policy, we usually want to know it. In a finite \ac{MDP}, there must always be at least one policy which gives this highest value of all possible policies.
  \item \textbf{Q-Value} The reward of taking action a in state s plus the value following a policy after that $Q^{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} V^{\pi}(s')$ is called Q-Value. When following the optimal policy, we simply write $Q^*(s,a)$. Usually, our Q functions are discounted similar to how we discount values. Note that we can define Values and Q-Values on each other: $V^{\pi}(s) = \mathbb{E}_{a \sim \pi} Q(s, a)$. If we approximate Q by a function approximator parametrized with $\theta$, we write $Q(s, a;\theta)$ or omit it in case it is clear. We sometimes call a learnt value estimation \textit{critic}.
  \item \textbf{Actor-Critic} Actor is just another name for a learnt policy and critic just another name for an estimated Q-Value or Value function. These two frequently appear together, if that happens we are using an actor-critic type algorithm.
